---
title: "Prostate Cancer: Stepwise AIC and BIC"
author: "Ruichen Tang"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Objective

Compare the full eight-predictor linear model with models selected using AIC and BIC, then evaluate all three specifications on the recommended held-out test set.

```{r packages}
library(MASS)
library(knitr)
```

## Data preparation

```{r data}
prostate_url <- "https://hastie.su.domains/ElemStatLearn/datasets/prostate.data"
prostate <- read.table(prostate_url, header = TRUE)

train_data <- subset(prostate, train == TRUE)
test_data <- subset(prostate, train == FALSE)
train_data$train <- NULL
test_data$train <- NULL

c(train_rows = nrow(train_data), test_rows = nrow(test_data))
```

## Full model and influence diagnostics

```{r full-model}
full_model <- lm(
  lpsa ~ lcavol + lweight + age + lbph + svi + lcp + gleason + pgg45,
  data = train_data
)

summary(full_model)
```

```{r diagnostics, fig.width=8, fig.height=7}
par(mfrow = c(2, 2))
plot(full_model)
par(mfrow = c(1, 1))

cook_distance <- cooks.distance(full_model)
data.frame(
  observation = which(cook_distance > 4 / length(cook_distance)),
  cooks_distance = cook_distance[cook_distance > 4 / length(cook_distance)]
)
```

## Stepwise selection

```{r selection}
aic_model <- stepAIC(full_model, direction = "both", trace = FALSE)
bic_model <- stepAIC(
  full_model,
  direction = "both",
  k = log(nrow(train_data)),
  trace = FALSE
)

selected_terms <- data.frame(
  model = c("Full", "AIC", "BIC"),
  predictors = c(
    paste(names(coef(full_model))[-1], collapse = ", "),
    paste(names(coef(aic_model))[-1], collapse = ", "),
    paste(names(coef(bic_model))[-1], collapse = ", ")
  )
)
kable(selected_terms)
```

## Held-out comparison

```{r evaluation}
models <- list(Full = full_model, AIC = aic_model, BIC = bic_model)

metrics <- do.call(rbind, lapply(names(models), function(model_name) {
  model <- models[[model_name]]
  predictions <- predict(model, newdata = test_data)
  data.frame(
    model = model_name,
    predictors = length(coef(model)) - 1,
    train_r_squared = summary(model)$r.squared,
    test_mse = mean((test_data$lpsa - predictions)^2),
    test_rmse = sqrt(mean((test_data$lpsa - predictions)^2))
  )
}))

kable(metrics, digits = 3)
```

The comparison illustrates the central tradeoff: AIC retains a relatively rich specification, whereas BIC applies a stronger complexity penalty and selects a two-variable model. In the original course run, the BIC specification achieved the lowest test MSE despite its lower training R².

