---
title: "Wisconsin Breast Cancer: Selection Stability and Ridge Logistic Regression"
author: "Ruichen Tang"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(527)
```

## Objective

Measure how correlated predictors affect stepwise selection stability, then fit a ridge logistic model that shrinks all coefficients jointly.

```{r packages}
library(MASS)
library(glmnet)
library(caret)
library(corrplot)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

## Data preparation

```{r data}
data_path <- file.path("..", "data", "breast_cancer_wisconsin.csv")
breast_cancer <- read.csv(data_path, check.names = FALSE)

breast_cancer$diagnosis <- factor(
  breast_cancer$diagnosis,
  levels = c("B", "M")
)
breast_cancer$id <- NULL
breast_cancer$X <- NULL

set.seed(527)
train_index <- createDataPartition(
  breast_cancer$diagnosis,
  p = 0.60,
  list = FALSE
)

train_data <- breast_cancer[train_index, ]
test_data <- breast_cancer[-train_index, ]

data.frame(
  split = c("Train", "Test"),
  rows = c(nrow(train_data), nrow(test_data)),
  malignant_rate = c(
    mean(train_data$diagnosis == "M"),
    mean(test_data$diagnosis == "M")
  )
) |> kable(digits = 3)
```

## Correlation structure

```{r correlation, fig.width=10, fig.height=9}
x_train <- model.matrix(diagnosis ~ ., data = train_data)[, -1]
y_train <- train_data$diagnosis
x_test <- model.matrix(diagnosis ~ ., data = test_data)[, -1]
y_test <- test_data$diagnosis

corrplot(
  cor(x_train),
  method = "color",
  type = "upper",
  tl.cex = 0.55,
  tl.col = "black",
  title = "Predictor Correlation Matrix",
  mar = c(0, 0, 2, 0)
)
```

## Ridge logistic regression

```{r ridge, fig.width=7, fig.height=5}
set.seed(527)
ridge_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 0,
  nfolds = 10,
  standardize = TRUE
)

plot(ridge_cv)

ridge_class <- predict(
  ridge_cv,
  newx = x_test,
  s = "lambda.min",
  type = "class"
)

ridge_accuracy <- mean(as.vector(ridge_class) == as.character(y_test))

data.frame(
  lambda_min = ridge_cv$lambda.min,
  lambda_1se = ridge_cv$lambda.1se,
  test_accuracy = ridge_accuracy
) |> kable(digits = 4)
```

## Bootstrap stability of stepwise selection

The calculation records failed bootstrap fits as missing rather than incorrectly treating every variable as unselected.

```{r bootstrap}
set.seed(527)
n_boot <- 100
variables <- setdiff(names(train_data), "diagnosis")

aic_selected <- matrix(
  NA_real_, nrow = n_boot, ncol = length(variables),
  dimnames = list(NULL, variables)
)
bic_selected <- aic_selected

for (iteration in seq_len(n_boot)) {
  bootstrap_index <- sample(seq_len(nrow(train_data)), replace = TRUE)
  bootstrap_data <- train_data[bootstrap_index, ]

  tryCatch({
    full_logistic <- glm(
      diagnosis ~ ., data = bootstrap_data, family = "binomial"
    )

    aic_fit <- suppressWarnings(
      stepAIC(full_logistic, direction = "both", trace = FALSE)
    )
    bic_fit <- suppressWarnings(
      stepAIC(
        full_logistic,
        direction = "both",
        k = log(nrow(bootstrap_data)),
        trace = FALSE
      )
    )

    aic_selected[iteration, ] <- 0
    bic_selected[iteration, ] <- 0
    aic_selected[iteration, variables %in% names(coef(aic_fit))] <- 1
    bic_selected[iteration, variables %in% names(coef(bic_fit))] <- 1
  }, error = function(error) NULL)
}

selection_frequency <- data.frame(
  variable = variables,
  AIC = colMeans(aic_selected, na.rm = TRUE) * 100,
  BIC = colMeans(bic_selected, na.rm = TRUE) * 100
) |>
  arrange(desc(AIC)) |>
  pivot_longer(c(AIC, BIC), names_to = "method", values_to = "frequency")
```

```{r stability-plot, fig.width=11, fig.height=6}
ggplot(
  selection_frequency,
  aes(x = reorder(variable, -frequency), y = frequency, fill = method)
) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(AIC = "#2166AC", BIC = "#B2182B")) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    title = "Bootstrap Selection Frequency",
    x = NULL,
    y = "Selection frequency (%)",
    fill = "Criterion"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 55, hjust = 1, size = 7),
    legend.position = "bottom"
  )
```

Correlated predictors can substitute for one another across resamples, so unstable inclusion should not be interpreted as proof that a feature is unimportant. Ridge addresses a different objective: it stabilizes prediction through continuous shrinkage rather than discrete selection.

