---
title: "METABRIC: Ridge versus Lasso"
author: "Angel Li"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(527)
```

## Objective

Compare ridge and lasso for predicting Nottingham Prognostic Index (NPI) in a high-dimensional setting, using the same train/test split and the same cross-validation folds for both methods.

```{r packages}
library(glmnet)
library(caret)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

## Data preparation

```{r data}
data_path <- file.path("..", "data", "METABRIC_RNA_Mutation.csv")

if (!file.exists(data_path)) {
  stop(
    "METABRIC_RNA_Mutation.csv was not found. Follow data/README.md, ",
    "then place the file in the data directory."
  )
}

raw_data <- read.csv(data_path, check.names = FALSE)

# Reproduce the course project's complete-case filtering, then retain numeric
# modeling fields. This produces the same candidate-predictor definition used
# in the final report.
model_data <- raw_data |>
  mutate(across(where(is.character), ~ na_if(.x, ""))) |>
  drop_na() |>
  select(where(is.numeric)) |>
  select(
    -any_of(c(
      "patient_id", "cohort", "tumor_size",
      "lymph_nodes_examined_positive", "neoplasm_histologic_grade"
    ))
  )

stopifnot("nottingham_prognostic_index" %in% names(model_data))

set.seed(527)
train_index <- createDataPartition(
  model_data$nottingham_prognostic_index,
  p = 0.70,
  list = FALSE
)

training <- model_data[train_index, ]
testing <- model_data[-train_index, ]

x_train <- training |>
  select(-nottingham_prognostic_index) |>
  as.matrix()
y_train <- training$nottingham_prognostic_index
x_test <- testing |>
  select(-nottingham_prognostic_index) |>
  as.matrix()
y_test <- testing$nottingham_prognostic_index

data.frame(
  complete_cases = nrow(model_data),
  candidate_predictors = ncol(x_train),
  training_rows = nrow(x_train),
  testing_rows = nrow(x_test)
) |> kable()
```

## Cross-validation and model fitting

```{r models, fig.width=10, fig.height=5}
set.seed(527)
fold_id <- sample(rep(seq_len(10), length.out = nrow(x_train)))

ridge_cv <- cv.glmnet(
  x_train, y_train,
  alpha = 0,
  foldid = fold_id,
  standardize = TRUE
)
lasso_cv <- cv.glmnet(
  x_train, y_train,
  alpha = 1,
  foldid = fold_id,
  standardize = TRUE
)

par(mfrow = c(1, 2))
plot(ridge_cv, main = "Ridge: Cross-Validation")
plot(lasso_cv, main = "Lasso: Cross-Validation")
par(mfrow = c(1, 1))

ridge_model <- glmnet(
  x_train, y_train,
  alpha = 0,
  lambda = ridge_cv$lambda.min,
  standardize = TRUE
)
lasso_model <- glmnet(
  x_train, y_train,
  alpha = 1,
  lambda = lasso_cv$lambda.min,
  standardize = TRUE
)
```

## Held-out performance

```{r evaluation}
score_model <- function(model, x_train, y_train, x_test, y_test) {
  train_prediction <- as.vector(predict(model, newx = x_train))
  test_prediction <- as.vector(predict(model, newx = x_test))

  data.frame(
    train_rmse = sqrt(mean((y_train - train_prediction)^2)),
    test_rmse = sqrt(mean((y_test - test_prediction)^2)),
    train_r_squared = 1 - sum((y_train - train_prediction)^2) /
      sum((y_train - mean(y_train))^2),
    test_r_squared = 1 - sum((y_test - test_prediction)^2) /
      sum((y_test - mean(y_test))^2),
    generalization_gap_mse =
      mean((y_test - test_prediction)^2) -
      mean((y_train - train_prediction)^2)
  )
}

metrics <- bind_rows(
  Ridge = score_model(ridge_model, x_train, y_train, x_test, y_test),
  Lasso = score_model(lasso_model, x_train, y_train, x_test, y_test),
  .id = "model"
)

kable(metrics, digits = 3)
```

The train-test MSE gap is reported only as a generalization diagnostic. It is not a formal estimate of variance, and training MSE is not treated as an estimate of squared bias.

## Sparsity and selected predictors

```{r sparsity}
ridge_coefficients <- as.matrix(coef(ridge_model))
lasso_coefficients <- as.matrix(coef(lasso_model))

complexity <- data.frame(
  model = c("Ridge", "Lasso"),
  nonzero_predictors = c(
    sum(ridge_coefficients[-1, 1] != 0),
    sum(lasso_coefficients[-1, 1] != 0)
  )
)
kable(complexity)

selected_lasso <- data.frame(
  predictor = rownames(lasso_coefficients)[-1],
  coefficient = lasso_coefficients[-1, 1]
) |>
  filter(coefficient != 0) |>
  arrange(desc(abs(coefficient)))

kable(head(selected_lasso, 20), digits = 4)
```

```{r metric-plot, fig.width=9, fig.height=5}
rmse_plot_data <- metrics |>
  select(model, train_rmse, test_rmse) |>
  pivot_longer(-model, names_to = "split", values_to = "rmse")

ggplot(rmse_plot_data, aes(split, rmse, fill = model)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = round(rmse, 3)),
    position = position_dodge(width = 0.7),
    vjust = -0.35
  ) +
  scale_fill_manual(values = c(Ridge = "#3B82F6", Lasso = "#BE3455")) +
  labs(title = "RMSE by Model and Split", x = NULL, y = "RMSE") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r predicted-actual, fig.width=10, fig.height=5}
prediction_data <- bind_rows(
  data.frame(
    actual = y_test,
    predicted = as.vector(predict(ridge_model, newx = x_test)),
    model = "Ridge"
  ),
  data.frame(
    actual = y_test,
    predicted = as.vector(predict(lasso_model, newx = x_test)),
    model = "Lasso"
  )
)

ggplot(prediction_data, aes(actual, predicted, color = model)) +
  geom_point(alpha = 0.55) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~ model) +
  scale_color_manual(values = c(Ridge = "#3B82F6", Lasso = "#BE3455")) +
  labs(
    title = "Predicted versus Actual Nottingham Prognostic Index",
    x = "Actual NPI",
    y = "Predicted NPI"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

Lasso's non-zero coefficients identify predictive associations under this modeling specification. They should not be interpreted as causal effects or clinically validated biomarkers.

